Rotation Periods in Low-Mass Binary Stars: The Impact of Tidal Torques and Magnetic Braking

### 2018-11-18 Paper/sync.tex
Updates from ShareLaTeX

\title{Rotation Periods in Low-Mass Binary Stars: The Impact of Tidal Torques}

@@

talk about how most people have focused on circularization, sync is hard to measure, we focus on low-mass stars with convective envelopes where the eqtide model is applicable, not higher mass stars with radiative envelopes where dynamical tide effects likely dominate (e.g. stuff lurie talked about)

Angular momentum evolution in low-mass (M$\lsim 1$ M$_{\odot}$) short-period (P$_{orb} \lsim 10$ d) stellar binaries is dominated by tides.  Tidal torques drive secular changes in the binary orbit and stellar spins, eventually circularizing the orbit and synchronizing the stellar spins in the long-term \citep{Counselman1973}. Orbital circularization in ubiquitous in short-period binaries, owing to the tidal torque's strong semi-major axis dependence, with both theoretical \citep[e.g.][]{Zahn1989} and observational \citep[e.g.][]{Meibom2005,Mazeh2008,Lurie2017} studies finding that most binaries with P$_{orb} \lsim 10$ d are circularized. For short-period binaries, tidal torques work quickly on ${\sim}100$ Myr timescales, as \citet{Zahn1989} finds that the orbit of solar twin binaries circularizes during the stellar pre-main sequence.  Observations by \citet{Meibom2005} support this picture as they find short-period binaries in the ${\sim}150$ Myr old cluster, M35, tend to have circular orbits.

Tides impart a significant signature in the long-term angular momentum evolution for binary stars, especially so for the stellar spins. Tidal-locking occurs much earlier than orbital circularization as the tidal-locking timescale is estimated to be $2-3$ orders of magnitude less than the circularization timescale since the stellar spins harbor much less angular momentum than the orbit \citep{Zahn1989,Witte2002,Mazeh2008}. Tidal-locking is expected for binaries with P$_{orb} \lsim$ 20 d \citep[e.g.][]{Meibom2006,Mazeh2008,Zahn2008,Meibom2015}. A large sample of $816$ rotation period measurements derived using starspot modulations on primary stars of \kepler eclipsing binaries by \citet{Lurie2017} confirm that most short-period binaries are tidally-locked, but also find tentative evidence for tidal-locking in binaries with P$_{orb}$ up to 45 d, their detection limit. Furthermore, \citet{Lurie2017} finds a subsynchrnous population of short-period binaries clustered near P$_{orb}/$P$_{rot}{\sim} 0.9$, in defiance of the expectation of tidal-locking.  Observations by the extended \kepler mission \citep[K2,][]{Howell2014} and the Transiting Exoplanet Survey Satellite \citep[TESS, ][]{Ricker2014,Sullivan2015} will contain low-mass, short-period eclipsing binaries, permitting additional tests for models of stellar rotation evolution due to tidal torques.

Rotation period distributions measured in clusters and field stars are likely contaminated by unresolved binaries, given that roughly half of Sun-like stars are in stellar binaries \citep{Raghavan2010,Duchene2013}, and that binaries are difficult to resolve in photometric surveys due to low transit probabilities. In the \kepler field, for example, \citet{Simonian2018} recently found that most rapid rotators (P$_{rot} \lsim 7.5$ d) are likely non-eclipsing, tidally-synchronized short-period photometric binaries, indicating that tidal torques in binaries can significantly impact observed P${rot}$ distributions.  Even if stellar binaries have not yet tidally-locked, tidal torques modify stellar rotation periods away from the long-term spin-down due to magnetic braking predicted for single stars \citep{Dunn1961,Skumanich1972,Barnes2003}. Ages inferred from rotation periods of stars in unresolved binaries using gyrochronology, a method that link the long-term spin-down of low-mass stars to their ages, \citep{Skumanich1972,Barnes2003,Barnes2007,Mamajek2008,Barnes2010}, will likely be incorrect, however no previous study has quantified this effect.  Moreover, no previous study has quantified how tidal-locking proceeds in stellar binaries using a realistic treatment of stellar evolution and magnetic braking over the full main sequence lifetimes of low-mass stars.

Here, we examine how tidal torques, coupled with a modern stellar evolution and magnetic braking models, shape the angular momentum evolution of low-mass binary stars.  We investigate under what conditions tidal-locking occurs, and what the typical tidal-locking timescale is as a function of stellar mass and tidal dissipation parameters for two widely-used equilibrium tidal models.  We show how tidal torques in binaries cause stellar rotation periods to not strongly correlate with age, causing the predictions of gyrochronlogy models to fail in such systems.  We describe our model in $\S$~\ref{sec:methods} and our simulation procedure in $\S$~\ref{sec:simulations}.  We discuss how our model impacts rotation period distributions and the implication for gyrochronology methods in $\S$~\ref{sec:gyro}, apply our model to the \kepler field in $\S$~\ref{sec:kepler}, and discuss our results' implications in $\S$~\ref{sec:discussion}.

@@

 Equilibrium tidal models, first introduced by \citep{Darwin1880}, track the secular evolution of an orbiter's semi-major axis, $a$, eccentricity, $e$, and the rotation rates, $\omega$, and obliquities $\psi$, of the both gravitating bodies due to tidal torques. Equilibrium tidal models assume that tidally-interacting bodies raise tidal bulges on their companions that remain offset from the line connecting the both bodies' centers of mass due to friction within each body.  The tidal bulges cause torques that drive the exchange of angular momentum between the orbit and both bodies' spins. Equilibrium tidal models are linear since they assume that the tidal waves that comprise the tidal bulge raised on a body are uncoupled. Under these assumptions, the tidal evolution is analogous to a driven, damped harmonic oscillator \citep{Greenberg2009}.  We refer the reader to \citet{Barnes2017} for an in-depth discussion of the assumptions and limitations of equilibrium tidal models.  Although simple, equilibrium tidal models have been used to robustly model the secular orbital and rotation evolution of both Solar System bodies and exoplanets \citet[e.g.][]{Goldreich1966,Jackson2009,Leconte2010,Heller2011,Barnes2013,Barnes2017} and stellar binaries \citep[e.g.][]{Zahn1989,Zahn2008,Khaliullin2011,Repetto2014,Fleming2018}.  Here, we consider two equilibrium tidal models to study the secular spin-orbital evolution of low-mass stellar binaries.

@@

All code used to run simulations and generate figures is available online\footnote{\href{https://github.com/dflemin3/sync}{https://github.com/dflemin3/sync}.} with accompanying descriptions and documentation.

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/eqPer.pdf}
   \caption{Caption.}%
    \label{fig:eqPer}%
\end{figure}

@@

\begin{figure*}[ht]
	\includegraphics[width=\textwidth]{../Plots/qTauPorbRatioHist.pdf}
   \caption{Caption}%
    \label{fig:qTauLock}%
\end{figure*}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/porbProtPorbQHist.pdf}
   \caption{Caption}%
    \label{fig:qmap}%
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/porbProtPorbTauHist.pdf}
   \caption{Caption}%
    \label{fig:taumap}%
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/subsync.pdf}
   \caption{Caption}%
    \label{fig:subsync}%
\end{figure}

>>>>

\title{Rotation Periods in Low-Mass Binary Stars: The Impact of Tidal Torques and Magnetic Braking}

@@

Angular momentum evolution in low-mass (M$\lsim 1$ M$_{\odot}$) short-period (P$_{orb} \lsim 10$ d) stellar binaries is dominated by tides.  Tidal torques drive secular changes in the binary orbit and stellar spins, eventually circularizing the orbit and synchronizing the stellar spins in the long-term \citep{Counselman1973}. Orbital circularization is ubiquitous in short-period binaries, owing to the tidal torque's strong semi-major axis dependence, with both theoretical \citep[e.g.][]{Zahn1989} and observational \citep[e.g.][]{Meibom2005,Mazeh2008,Lurie2017} studies finding that most binaries with P$_{orb} \lsim 10$ d are circularized. For short-period binaries, tidal torques work quickly on ${\sim}100$ Myr timescales, as \citet{Zahn1989} finds that the orbit of solar twin binaries circularizes during the stellar pre-main sequence.  Observations by \citet{Meibom2005} support this picture as they find short-period binaries in the ${\sim}150$ Myr old cluster M35 tend to have circular orbits.

Tides impart a significant signature in the long-term angular momentum evolution for binary stars, especially so for the stellar spins. Tidal torques drive stellar spins towards the tidally-locked state in which the stellar rotation periods (P$_{rot}$) are equal to the equilibrium rotation period (P$_{eq}$) predicted by tidal models, with a familiar example of this effect being spin-orbit synchronization where P$_{rot} = $ P$_{eq} = $ P$_{orb}$.  Tidal-locking occurs much earlier than orbital circularization as the tidal-locking timescale is estimated to be $2-3$ orders of magnitude less than the circularization timescale \citep{Zahn1989,Witte2002,Mazeh2008}, and as a result, tidal-locking is expected for binaries with P$_{orb} \lsim$ 20 d \citep[e.g.][]{Meibom2006,Mazeh2008,Zahn2008,Meibom2015}.

A large, homogenous sample of $816$ rotation period measurements derived using starspot modulations on low-mass \kepler eclipsing binary stars by \citet{Lurie2017} confirm that most short-period binaries are tidally-locked and also show that the tidal torques can influence stellar spins in binaries with P$_{orb}$ up to 45 d, their detection limit. Futhermore, \citet{Lurie2017} shows that a substantial a population of subsynchronous short-period binaries exist in the \kepler field, clustered near P$_{orb}/$P$_{rot}{\sim} 0.9$, in defiance of the expectation of tidal locking.  These findings suggest that that synchronization for short P$_{orb}$ binaries is not the sole outcome of tidal interactions in low-mass stellar binaries, but rather that impact of tidal effects should be reconsidered for longer P$_{orb}$ and to examine if our current understanding of tidal torques is consistent with short P$_{orb}$ subsynchronous rotators.  Both the extended \kepler mission \citep[K2,][]{Howell2014} and the Transiting Exoplanet Survey Satellite \citep[TESS, ][]{Ricker2014,Sullivan2015} will detect additional low-mass, short-period eclipsing binaries, permitting additional tests for models of stellar angular momentum evolution subject to tidal torques.

Understanding angular momentum evolution in low-mass stellar binaries is paramount as rotation period distributions measured in clusters and field stars are likely contaminated by unresolved binaries, given that roughly half of Sun-like stars are in stellar binaries \citep{Raghavan2010,Duchene2013}, and that binaries are difficult to resolve in photometric surveys due to low transit probabilities. In the \kepler field, for example, \citet{Simonian2018} recently found that most rapid rotators (P$_{rot} \lsim 7.5$ d) are likely non-eclipsing, tidally-synchronized short-period photometric binaries, indicating that tidal torques in binaries can significantly impact observed P${rot}$ distributions.  Even if stellar binaries have not yet tidally-locked, tidal torques modify stellar rotation periods towards the tidally-locked state, away from the long-term spin-down due to magnetic braking predicted for single stars \citep{Dunn1961,Skumanich1972,Barnes2003}, impacting a contaminating signal that current models do not account for. Moreover, any ages inferred from rotation periods of stars in unresolved binaries using gyrochronology, a method that link the long-term spin-down of low-mass stars to their ages, \citep{Skumanich1972,Barnes2003,Barnes2007,Mamajek2008,Barnes2010}, will likely be incorrect owing to the influence of tidal torques. No previous study has quantified this effect.

Here, we examine angular momentum evolution over the full main sequence lifetimes of low-mass stellar binaries using a realistic treatment of stellar evolution, magnetic braking, and tidal torques. We investigate under what conditions tidal-locking occurs, and how tidal torques influence rotation in stellar binaries as a function of binary orbital period and tidal dissipation parameters for two widely-used equilibrium tidal models.  We show how tidal torques can impact stellar rotation rate in binaries out to P$_{orb} = 100$ d, causing stellar rotation periods to not strongly correlate with age, making the predictions of gyrochronlogy models to fail in such systems.  We describe our model in $\S$~\ref{sec:methods} and our simulation procedure in $\S$~\ref{sec:simulations}.  We discuss how our model impacts rotation period distributions and the implication for gyrochronology methods in $\S$~\ref{sec:gyro}, apply our model to the \kepler field in $\S$~\ref{sec:kepler}, and discuss our results' implications in $\S$~\ref{sec:discussion}.

@@

 Equilibrium tidal models, first introduced by \citep{Darwin1880}, track the secular evolution of an orbiter's semi-major axis, $a$, eccentricity, $e$, and the rotation rates, $\omega$, and obliquities $\psi$, of the both gravitating bodies due to tidal torques. Equilibrium tidal models assume that tidally-interacting bodies raise tidal bulges on their companions that remain offset from the line connecting the both bodies' centers of mass due to friction within each body.  The tidal bulges cause torques that drive the exchange of angular momentum between the orbit and both bodies' spins. Equilibrium tidal models are linear since they assume that the tidal waves that comprise the tidal bulge raised on a body are uncoupled. Under these assumptions, the tidal evolution is analogous to a driven, damped harmonic oscillator \citep{Greenberg2009}. We refer the reader to \citet{Barnes2017} for an in-depth discussion of the assumptions and limitations of equilibrium tidal models.  Although simple, equilibrium tidal models have been used to robustly model the secular orbital and rotation evolution of both Solar System bodies and exoplanets \citet[e.g.][]{Goldreich1966,Jackson2009,Leconte2010,Heller2011,Barnes2013,Barnes2017} and stellar binaries \citep[e.g.][]{Zahn1989,Zahn2008,Khaliullin2011,Repetto2014,Fleming2018}. For stars, equilibrium tidal models assume that tidal forces dissipate energy in convective regions via viscous turbulence \citep[see][]{Zahn2008}, corresponding tidal dissipation in the outer convective layers of low-mass stars. Here, we consider two equilibrium tidal models to study the secular spin-orbital evolution of low-mass stellar binaries.

@@

We also perform a smaller subset of simulations to illustrate the behaviour of our coupled model and describe their initial conditions as we introduce them. All code used to run simulations and generate figures is available online\footnote{\href{https://github.com/dflemin3/sync}{https://github.com/dflemin3/sync}.} with accompanying descriptions and documentation.

@@

\begin{figure*}
	\includegraphics[width=\textwidth]{../Plots/eqPerTwoPanel.pdf}
   \caption{Evolution of stellar P$_{rot}$, normalized by P$_{eq}$ (see Eqn.~(\ref{eqn:cpl:eqPer}) and Eqn.~(\ref{eqn:ctl:eqPer}), for initial circular binary orbits according to the CPL (solid) and CTL (dashed) models with $Q = 10^6$ and $\tau = 0.1$ s, respectively. Left: P$_{rot}/$P$_{eq}$ for stars with P$_{orb}$ ranging from 5 d to 60 d. The black dotted line indicates that tidally-locked state where P$_{rot} = $ P$_{eq}$.  For P$_{orb} \leq 10$ d, both the CPL and CTL models predict that the binaries lock into synchronous rotation.  For all P$_{orb}$, all CPL models tidally-lock into synchronous rotation whereas the CTL model predicts subsynchronous (P$_{rot} > $ P$_{eq}$) rotation. In the CTL cases, the tidal torques are weaker than the magnetic braking torque, causing the stars to continue to spin down, preventing the systems from tidally-locking.  Tidal torques strengthen for P$_{rot}$ far from P$_{eq}$, and magnetic braking torques weaken for long P$_{rot}$ \citep{Kawaler1988,Matt2015}, causing tidal torques to overpower magnetic braking, driving the stellar rotation back towards the tidally-locked state.  Right: Net P$_{rot}$ derivative due to stellar evolution, tidal torques, and magnetic braking.  We truncate each curve when the respective system tidally-locks.  Initially, all systems spin-up due to stellar contraction, except for the P$_{orb} = 5$ d case where tides overpower other torques, rapidly driving the stars towards synchronization. Once the stars reach the main sequence at ${\sim} 60$ Myr, tidal torques and magnetic braking combine to spin down the stars towards the tidally-locked state. For stars at long P$_{orb}$ that do not tidally-lock, after having spun-down to long P$_{rot}$, magnetic braking weakens, allowing tidal torques to sheppard P$_{rot}$ towards P$_{eq}$ as the P$_{rot}$ derivatives stay near 0, but with a slight negative value.}%
    \label{fig:eqPer}%
\end{figure*}

@@

\begin{figure*}[ht]
	\includegraphics[width=\textwidth]{../Plots/qTauPorbRatioHist.pdf}
   \caption{Median P$_{orb}/$P$_{rot}$ for the primary star at the end of the simulation according to the CPL (left) and CTL (right) models binned by log$_{10}(Q)$ and log$_{10}(\tau)$, respectively, and P$_{orb}$.  The CPL model predicts that stellar binaries can tidally-lock out to P$_{orb} = 30$ d for most values of $Q$, with some locking all the way up to P$_{orb} = 100$ d for small $Q < 10^5$, substantially longer than what is expected.  For P$_{orb} < 10$ d, the CPL model predicts that most systems will tidally-lock into synchronous rotation, regardless of $Q$.  The CTL model, however, requires stars to have $\tau \geq 0.1$ s in order for stellar binaries with P$_{orb} < 10$ d to likely tidally-lock.  Both models exhibit supersynchronous rotation (blue regions, P$_{rot} < $P$_{orb}$) for P$_{orb} > 80$d, that do not in general reflect tidally-locking into a supersynchronous rotation, but rather at these long P$_{orb}$, the combination of weak tidal torques and magnetic braking have not spin down the star enough for P$_{rot}$ to be close to P$_{orb}$ after 7 Gyr of evolution. For large $Q$ ($Q > 10^{6.5}$) and small $\tau$ ($\tau < 0.1$ s), weak tidal torques cannot prevent magnetic braking from spining down stars past the tidally-locked state, producing a population of subsynchronous (red regions, P$_{rot} > $P$_{orb}$) rotators.}%
    \label{fig:qTauLock}%
\end{figure*}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/porbProtPorbQHist.pdf}
   \caption{Median log$_{10}(Q)$ of primary stars binned by P$_{orb}$ and P$_{orb}/$P$_{rot}$.  In each bin, we compute the median, $16\%$, and $84\%$ percentiles for log$_{10}(Q)$. For all P$_{orb}$, synchronous and supersynchronous rotators have systematically low $Q$s as strong tidal torques are required to tidally-lock the systems. Note that for P$_{orb} < 10$ d, there are no rotators with $1.0 \leq$ P$_{orb}/$P$_{rot} \leq 1.5$ as strong tidal torques always tidally-lock the system, forcing P$_{rot} =$ P$_{eq}$, which can only take two values under the CPL model (see Eqn.~(\ref{eqn:cpl:eqPer})). Subsynchronous rotators have large $Q$s, and hence weak tidal torques, that are dominated by magnetic braking that spin-down the stars past the tidally-locked state.  For P$_{orb} < 70$ d, no stars have P$_{orb}/$P$_{rot} > 1.5$ as in the CPL model, supersynchronous rotators can only tidally-lock into a 3:2 spin-orbit commensurability.  At longer P$_{orb}$, supersynchronous rotation arises from the inability of tidal torques and magnetic braking to spin-down stars enough to approach the tidally-locked state by the end of the simulation.}%
    \label{fig:qmap}%
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/porbProtPorbTauHist.pdf}
   \caption{Same format as Fig.~\ref{fig:qmap}, but for the CTL model. As under the CPL model, stars tidally-lock when subject to strong tidal torques, typically for P$_{orb} < 50$ d and $\tau \gsim 0.25$ s.  For P$_{orb} < 50$ d, subsynchronous rotation occurs for stars with low $tau < 0.1$ s as magnetic braking dominates over tidal torques, spinning down the star.  For P$_{orb} > 50$ d, tides do not in general dominate the evolution as the diagonal sequence with a median $\tau \approx 0.1$ s has upper and lower limits that encompass most of the range we considered for $\tau$.  The shape of this diagonal region arises from the combination of magnetic braking and our flat initial P$_{orb}$ distribution. We do not observe P$_{orb}/$P$_{rot} \gsim 1.5$ as we only consider eccentricities up to $e = 0.3$, limiting how rapid supersynchronous systems can rotate according to Eqn.~(\ref{eqn:ctl:eqPer}).}%
    \label{fig:taumap}%
\end{figure}

@@

Median subsync tidal tau: 0.0790922272
Median supersync tidal tau: 0.12508653609999998
Median subsync ecc: 0.1288805
Median supersync ecc: 0.151426
Fraction with Porb/Prot in [0.92,1.2] for Porb < 10d: 0.6126373626373627
Fraction with Porb/Prot in [0.84,0.92] for Porb < 10d: 0.11401098901098901

Although Lurie+2017 attributes differential rotation to the production of the
subsynchronous population, we find that coupled stellar-tidal evolution naturally
produces the population. Compare with Lurie+2017 sample for 2 $<$ Porb $<$ 10 days:
$72\%$ in [0.92,1.2] and $15\%$ in [0.84, 0.92].

\begin{figure}
	\includegraphics[width=0.45\textwidth]{../Plots/subsync.pdf}
   \caption{P$_{orb}/$P$_{rot}$ as a function for P$_{orb}$ simulated according to the CTL model (dots) compared with \kepler EBs from \citet{Lurie2017} (diamonds), colored by $e$. Our simulations are in excellent agreement with the observed orbital and rotational states of \kepler EBs. Our model predicts that all EBs with P$_{orb} < 3$ d are synchronized as is observed by \citet{Lurie2017}. We find that our model predicts a population of supersyncronous rotators on eccentric orbits and a population of subsyncronous rotators on nearly-circular orbits, both of which were observed by \citet{Lurie2017}. This excellent agreement between theory and observations validates our model and provides strong evidence that the CTL formalism can accurately model the tidal evolution of low-mass stellar binaries.}%
    \label{fig:subsync}%
\end{figure}
